---
title: "**Unlocking Profitable Insights: A Comprehensive Analysis of Sales, Profit, and Customer Behavior in the EU Superstore Data**"
author: "Zunaira Azam"
date: "2024-01-20"
output: html_document

---

## **Objectives**

Objective of the study are:


- Identify the top-selling region in the EU Superstore dataset.


- Identify top-selling product categories driving overall revenue.


- Examine key subcategories driving sales to gain insights into customer-favored product types.


- Use rolling window analysis to smooth out short-term fluctuations and reveal trends in metrics such as sales, profit, quantity, and discount by calculating their rolling averages over a set time window.


- Examine quarterly sales data, group by quarters, and create visualizations to reveal seasonal patterns and fluctuations in Sample EU Superstore's performance.


- Analyze the profit margin for each product.


- Analyze which products or categories respond better to discounts.

## **Key Questions**

1. What are the top categories and sub-categories in terms of sales and profit?


2. Are there any specific products that contribute significantly to profit?


3. What is the distribution of sales, quantity, discount, and profit?


4. How are sales and profit distributed across different regions, categories, and sub-categories?


5. Which region has the highest sales?


6. How do profit and sales vary across different regions?


7. How does the average sales vary across different ship modes?


8. Are there any products that can consistently contribute to high profits?


9. How does the application of discounts impact the sales performance of different products or categories?


10. How does the profit margin vary across different products?


### **Project Analysis**

```{r, echo=FALSE, results='hide'}
# Import dataset
library(readxl)
# Import dataset
Sample_EU_Superstore <- read_excel("C:/Users/Mg/Documents/Sample - EU Superstore.xlsx")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(plotly)
library(corrplot)
library(lubridate)
library(zoo)
```

```{r,echo=FALSE, results='hide'}
# Missing values summary
as.data.frame(summary(is.na(Sample_EU_Superstore)))
```

### **1. Summary**

```{r}
summary_df <- summary(Sample_EU_Superstore[c("Sales", "Quantity", "Discount", "Profit")])
summary_df <- as.data.frame(as.table(summary_df))
colnames(summary_df) <- c("Variable", "Statistic", "Value")

# Replace 'Variable' column values with the actual variable names
summary_df$Variable <- c("Sales", "Quantity", "Profit", "Discount")

# Assuming your summary output looks something like this
summary_df <- summary(Sample_EU_Superstore[c("Sales", "Quantity", "Discount", "Profit")])
summary_df <- as.data.frame(as.table(summary_df))
colnames(summary_df) <- c("Variable", "Statistic", "Value")

# Replace 'Variable' column values with the actual variable names
summary_df$Variable <- c("Sales", "Quantity", "Profit", "Discount")

```

### **2. Total Sales Across Regions**

```{r}
# Region-wise analysis
region_sales <- Sample_EU_Superstore %>% 
  group_by(Region) %>% 
  summarise(total_sales = sum(Sales))
as.data.frame(print(region_sales))
```
```{r}
# Bar plot for sales across regions with Plotly
region_sales_plot<- ggplot(region_sales, aes(x = Region, y = total_sales, fill = Region)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Sales Across Regions", x = "Region", y = "Total Sales") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# Convert ggplot to plotly
region_plot<-ggplotly(region_sales_plot)
region_plot
```

### **3. Proportions for Region Sales**

```{r}
#Proportions
region_sales_prop <- Sample_EU_Superstore %>% 
  group_by(Region) %>% 
  summarise(total_sales = sum(Sales))
# Calculate the percentage contribution
region_sales_prop$Percentage <- (region_sales$total_sales / sum(region_sales$total_sales)) * 100
# Print the results
print(region_sales_prop)
```
```{r}
# Stacked bar plot for percentage contribution of sales by region with Plotly
region_plot <- ggplotly(
  ggplot(region_sales_prop, aes(x = Region, y = Percentage, fill = Region)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = "Percentage Contribution of Sales by Region",
         x = "Region",
         y = "Percentage (%)") +
    theme_minimal()
)
region_plot
```

### **4. Category Wise Analysis**

```{r}
# Category-wise analysis
category_sales <- Sample_EU_Superstore %>% 
  group_by(Category) %>% 
  summarise(total_sales = sum(Sales)) 
  print(category_sales)
```
```{r}
# Convert total_sales to decimal format
category_sales$total_sales <- format(category_sales$total_sales, scientific = FALSE)
# Bar plot for sales across categories
category_sales_plot <- ggplot(category_sales, aes(x = Category, y = total_sales, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Sales Across Categories", x = "Category", y = "Total Sales") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# Convert ggplot to plotly
category_sales_plotly <- ggplotly(category_sales_plot)
# Display the plotly object
category_sales_plotly
```

### **5. Sub_Category Analysis**

```{r}
## Sub-category-wise analysis
sub_category_profit <- Sample_EU_Superstore %>% 
  group_by(`Sub-Category`) %>% 
  summarise(total_profit = sum(Profit))
print(sub_category_profit)
```
```{r}
sub_category_profit_plot <- ggplot(sub_category_profit, aes(x = `Sub-Category`, y = total_profit, fill = `Sub-Category`)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Profit Across Sub-categories", x = "Sub-category", y = "Total Profit") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
sub_category_profit_plotly <- ggplotly(sub_category_profit_plot)
sub_category_profit_plotly

```

### **6. Quarterly Sales Analysis**

```{r}
# Convert Order Date to Date format
Sample_EU_Superstore$`Order Date` <- as.Date(Sample_EU_Superstore$`Order Date`)

# Extract quarter from Order Date
Sample_EU_Superstore$Quarter <- quarter(Sample_EU_Superstore$`Order Date`, with_year = TRUE)

# Aggregate Quarterly Sales by Segments
quarterly_sales <- Sample_EU_Superstore %>%
  group_by(Segment, Quarter) %>%
  summarise(total_sales = sum(Sales))
print(quarterly_sales)
```
```{r}
# Plot Quarterly Sales by Segments
quarterly_sales_plot <- ggplot(quarterly_sales, aes(x = Quarter, y = total_sales, color = Segment)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  labs(title = "Quarterly Sales by Segments",
       x = "Quarter",
       y = "Total Sales",
       color = "Segments") +
  theme_minimal()
# Display the interactive plot using plotly
ggplotly(quarterly_sales_plot)

```

### **7. Yearly Sales Forecast** 

```{r}
# Convert Order Date to Date format
Sample_EU_Superstore$`Order Date` <- as.Date(Sample_EU_Superstore$`Order Date`)
# Define a function to compute the desired metric (e.g., rolling average) within the rolling window
rolling_window_function <- function(data) {
  # Replace 'Sales' with the desired metric (Sales, Profit, Quantity, Discount)
  result <- mean(data, na.rm = TRUE)
  return(result)
}
# Set the size of the rolling window (e.g., 3 months)
window_size <- 3
# Apply rolling window analysis to the entire dataset
rolling_window_results <- Sample_EU_Superstore %>%
  arrange(`Order Date`) %>%
  group_by(`Order Date`) %>%
  summarise(Sales = sum(Sales)) %>%
  mutate(rolling_sales = zoo::rollapply(Sales, window_size, rolling_window_function, align = "right", fill = NA, partial = TRUE))
print(rolling_window_results) # Display the results
```
```{r}
print(rolling_window_results)
# Line chart for rolling sales with enhanced readability
plot <- ggplot(rolling_window_results, aes(x = `Order Date`, y = rolling_sales)) +
  geom_line(color = "lightblue", linewidth = 1, linetype = "solid") +
  geom_smooth(method = "loess", se = FALSE, color = "green", linetype = "dashed") +
  labs(title = "Rolling Window Analysis - Sales Over Time",
       x = "Order Date",
       y = "Rolling Sales",
       caption = "Dashed line represents a smoothed trend") +
  theme_minimal() +
  theme(legend.position = "none") +
  annotate("text", x = max(rolling_window_results$`Order Date`), y = max(rolling_window_results$rolling_sales), 
           label = "Smoothed Trend", hjust = 1.2, vjust = 0.5, color = "orange") +
  geom_point(size = 2, color = "pink") 
rolling_window_plot <- ggplotly(plot); rolling_window_plot
```

### **8. Yearly Sales and Profit**

```{r}
# Convert 'Order Date' to a Date type
Sample_EU_Superstore$`Order Date` <- as.Date(Sample_EU_Superstore$`Order Date`)

# Extract year from 'Order Date'
Sample_EU_Superstore <- Sample_EU_Superstore %>%
  mutate(Year = year(`Order Date`))

# Group by year and calculate total sales and profit
yearly_summary <- Sample_EU_Superstore %>%
  group_by(Year) %>%
  summarise(TotalSales = sum(Sales),
            TotalProfit = sum(Profit))
print(yearly_summary)
```
```{r}
# Visualize the results with a line chart
ggplotly(ggplot(yearly_summary, aes(x = factor(Year), y = TotalSales, group = 1)) +
  geom_line(aes(color = "Total Sales"), size = 1) +
  geom_line(aes(y = TotalProfit, color = "Total Profit"), size = 1) +
  ggtitle("Yearly Sales and Profit") +
  xlab("Year") +
  ylab("Amount") +
  scale_color_manual(values = c("Total Sales" = "#4e79a7", "Total Profit" = "#e15759")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white"),
    legend.position = "top"
  )
)

```

### **9. TOP 10 Products by Total Profit**

```{r}
###Product Analysis
# Group by product and calculate total sales, profit, and profit margin
product_summary <- Sample_EU_Superstore %>%
  group_by(`Product ID`, `Product Name`) %>%
  summarise(TotalSales = sum(Sales),
            TotalProfit = sum(Profit),
            ProfitMargin = sum(Profit) / sum(Sales) * 100) %>%
  arrange(desc(TotalProfit))

# Select the top N products (adjust N as needed)
top_products <- head(product_summary, 10)

# Visualize the results with rotated x-axis labels
ggplotly(ggplot(top_products, aes(x = reorder(`Product Name`, -TotalProfit), y = TotalProfit)) +
  geom_bar(stat = "identity",fill = "yellow") +
  ggtitle("Top 10 Products by Total Profit") +
  xlab("Product Name") +
  ylab("Total Profit") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate labels for better readability
)
```

### **10. Average Sales with Different Discounts**

```{r}
#Discount Analysis
# Assuming you have a column 'Discount' in your dataset
discount_analysis <- Sample_EU_Superstore %>%
  group_by(Category, `Product Name`, Discount) %>%
  summarise(AvgSales = mean(Sales))
discount_analysis
```

```{r}
# Visualize the results

plot <- ggplot(discount_analysis, aes(x = Discount, y = AvgSales, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Average Sales with Different Discounts") +
  xlab("Discount") +
  ylab("Average Sales") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Use ggplotly to convert to interactive plotly plot
plotly_plot <- ggplotly(plot)
plotly_plot
```

### **11. Top Profitable Products**

```{r}
#product analysis
# Calculate total profit for each product
product_profit <- Sample_EU_Superstore %>%
  group_by(`Product ID`, `Product Name`) %>%
  summarize(TotalProfit = sum(Profit))

# Identify top products with consistently high profits
top_profitable_products <- product_profit %>%
  filter(TotalProfit > quantile(TotalProfit, 0.75))  # Adjust the quantile threshold as needed
top_profitable_products
```

```{r}
# Visualize top profitable products
ggplotly(ggplot(top_profitable_products, aes(x = reorder(`Product Name`, -TotalProfit), y = TotalProfit)) +
  geom_bar(stat = "identity", fill = "pink") +
  labs(title = "Top Profitable Products",
       x = "Product Name",
       y = "Total Profit") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
)
```

### **12. Average Sales Across Different Ship Modes**

```{r}
###ship mode analysis: 
# Group the data by ship mode and calculate the average sales for each mode
ship_mode_sales <- Sample_EU_Superstore %>%
  group_by(`Ship Mode`) %>%
  summarise(AverageSales = mean(Sales, na.rm = TRUE))
# Print the result
print(ship_mode_sales)
```

```{r}
# Create a more attractive bar plot
ggplotly(ggplot(ship_mode_sales, aes(x = reorder(`Ship Mode`, -AverageSales), y = AverageSales, fill = `Ship Mode`)) +
  geom_bar(stat = "identity", size = 0.7) +
  labs(title = "Average Sales Across Different Ship Modes",
       x = "Ship Mode",
       y = "Average Sales") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="none") +
  scale_fill_brewer(palette = "Set3")  # You can choose a different color palette
)
```